---
title: "Tutorial 1"
author: "andy south"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This tutorial will be about you doing stuff. I will get you to try things probably before you understand fully what they are. Worry not.

1.  It doesn't matter if you make mistakes.
1.  We will come back to some of the concepts later.
1.  Questions are encouraged.
1.  This should at least give you a start to follow up on later.

The beauty (& sometimes otherwise) of R is that there are usually multiple ways of doing the same thing. Here I will show one way, this will help you learn some concepts but you may well find that alternatives are better suited to your needs.


### Install the R package for the course

```{r eval=FALSE, tidy=FALSE}
#first install devtools to enable install from GitHub 
install.packages("devtools")
library(devtools)
install_github('AndySouth/rmapteach')
```

Load the R package for the course
```{r eval=TRUE}
library("rmapteach")
```


### A) point plots with ggplot2 first go

```{r eval=TRUE}

# load the package
library("ggplot2")

# load some data on mammal sleep patterns from the package
data(msleep)

# to show the 'str'ucture of the data
str(msleep)


# first plot
ggplot(data = msleep, aes(x = bodywt, y = awake)) +
    geom_point()

# aes (for aesthetic) sets up links between the data and aspects of the plot
# + adds layers to the plot, in this case points
# bodywt and awake are columns within the data

# add an extra column to the aes
ggplot(data = msleep, aes(x = bodywt, y = awake, col = vore)) +
    geom_point()

# if you change col= to a column that is numeric it changes the legend
ggplot(data = msleep, aes(x = bodywt, y = awake, col = sleep_rem)) +
    geom_point()

```

### A1) Exercise
Have a quick look at the online ggplot2 documentation here : http://docs.ggplot2.org/current/ 

Try modifying a point plot with the msleep data.


### point plots, modifying axes, label points, divide into subplots

```{r eval=TRUE}

# we can log bodywt to space the points out more
ggplot(data = msleep, aes(x = log(bodywt), y = awake, col = vore)) +
    geom_point()

# label the points for data exploration
ggplot(data = msleep, aes(x = log(bodywt), y = awake, label = name)) +
    geom_point() + 
    geom_text()

# divide into subplots using facet_wrap()
ggplot(data = msleep, aes(x = log(bodywt), y = awake, label = name)) +
    geom_point() + 
    facet_wrap(~vore)

``` 





### 1 maps

Install and load the tmap package
```{r eval=FALSE}
install.packages("tmap")
```

```{r eval=TRUE}
library("tmap")
```

### 1.1 Getting started with polygon maps
```{r eval=TRUE}

# load data from tmap
data(Europe)
# set shapes and fill them
tm_shape(Europe) +
  tm_fill("purple")

# load data from tmap
data(World)
# set shapes and fill them
tm_shape(World) +
  tm_fill("green")

# check what data are associated with a map using 'str(*@data)', str stands for structure
str(World@data)
str(Europe@data)
# note that str(Europe) shows all the geometry information too and is not very useful here

# check the contents of Factor variables using 'levels'
levels(Europe$part)
# levels(Europe@data$part) gives the same output, you don't always need the @data to get at the variables 

# plot a subset of the map using '[which(*),]'
tm_shape(Europe[which(Europe$part=='Northern Europe'),]) +
  tm_fill("red")


# check the contents of numeric variables using 'hist'
hist(Europe$pop_est)

# fill polygons on a map based on data variables
# factor
tm_shape(Europe) +
  tm_fill("part")
# numeric
tm_shape(Europe) +
  tm_fill("pop_est")

# the previous commands used default colours, you can modify these with 'palette'
tm_shape(Europe) +
  tm_fill("pop_est", palette = "YlGnBu")
# to see available palettes 
RColorBrewer::display.brewer.all()

# to add a histogram to the legend
tm_shape(Europe) +
  tm_fill("pop_est", palette = "YlGnBu", legend.hist = TRUE)
  
# maps are stored as a SpatialPolygonsDataFrame from the package sp
# you can find this out by typing
class(Europe)

```

### 1.2 Exercise : can you plot different areas, different variables, different colours ...

```{r eval=FALSE}
# to see help use '?'
?tm_fill

# syntax is like ggplot2 (e.g. + to add layers)

# start with one of the code examples above and change something ...


```



### 2.1 Getting started with point maps
```{r eval=TRUE}

# load some tmap point data for metropolitan areas
data(metro)
# ?metro gives information about the data
# this is also an sp object with associated variables
class(metro)
str(metro@data)

# plot points as bubbles
tm_shape(metro) +
    tm_bubbles("pop2010", legend.size.show = FALSE)

# plot points on top of a European map
tm_shape(Europe) +
  tm_borders() +
tm_shape(metro) + 
  tm_bubbles("pop2010")

```

### 2.2 Exercise - can you cahnge one of the point maps ?




### 3.1 Getting started with raster maps

```{r eval=TRUE}

# load some data from tmap
data(land)

class(land)

str(land@data)

tm_shape(land) + 
    tm_raster("trees", breaks=seq(0, 100, by=20) )

tm_shape(land) + 
    tm_raster("trees", breaks=seq(0, 100, by=50) )

# categorical land cover
tm_shape(land) + 
    tm_raster("cover")

# using faceting

# facet by cover to get one map for each cover type
tm_shape(land) + 
    tm_raster("cover", legend.show = FALSE) +
    tm_facets("cover", free.coords=TRUE, drop.units=TRUE)   


# facet by country to get one cover map per country
tm_shape(land) + 
    tm_raster("cover") + # , legend.show = FALSE)
tm_shape(Europe[which(Europe$part=="Northern Europe"),]) +
    tm_borders() +
    tm_facets("name", free.coords = TRUE)

```


### 3.2 Exercise



### 4 Using your own data
You can use both your own geometry and/or attribute data in tmap and other mapping options.

```{r eval=TRUE}

# create a very simple data frame as an example
dF <- data.frame( country=c("Spain", "United Kingdom"),
                  weather=c("hot", "cold") )

# tmap can join this onto a map (other options are available too)
Eu_and_dat <- append_data(Europe, dF, key.shp="name", key.data="country")

tm_shape(Eu_and_dat) +
  tm_fill("weather", palette=c("red","blue"))

# Can you plot some other data by country ?
# remember the country names in the map are in Europe$name
# TODO me to add getting data from another package


```


### 5 Interactivity (with leaflet &/or tmap)


### 6 Putting it all together or experiment with other options
```{r eval=TRUE}
# to list datasets in the tmap package
data(package="tmap")



```

### 7 Going further
```{r eval=FALSE}
# other useful packages, resources

# raster package
library(raster)

# rmapshaper for simplifying polygon boundaries
library(rmapshaper)





```


###TODO
~ acknowledge Martijn
~ say I wrote an older package to do this kind of thing but tmap is newer & better (in almost all respects)


